\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-27 paper.qmd
tidied up fist image and code

```{r}
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)
library(tidyr)

library(Rtsne)
library(uwot)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
set.seed(20240110)

source("nldr_code.R", local = TRUE)
```

@@

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                                                 x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + 
  theme_linedraw() +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
        legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm')) 
```

```{r}
#| echo: false
#| fig-cap: "2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-umapauthor
#| fig-pos: H

umap_plot_pbmc 
```

>>>>

```{r include=FALSE}
# Set up chunk for for knitr
knitr::opts_chunk$set(
  fig.width = 5,
  fig.height = 5,
  fig.align = "center",
  out.width = "100%",
  code.line.numbers = FALSE,
  fig.retina = 4,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  dev.args = list(pointsize = 11)
)
```

```{r}
#| label: load-libraries
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)
library(tidyr)

library(Rtsne)
library(uwot)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
```

```{r}
#| label: plot-theme
theme_set(theme_linedraw() +
   theme(
     aspect.ratio = 1,
     plot.background = element_rect(fill = 'transparent', colour = NA),
     plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
     panel.background = element_rect(fill = 'transparent', 
                                     colour = NA),
     panel.grid.major = element_blank(), 
     panel.grid.minor = element_blank(), 
     axis.title.x = element_blank(), axis.title.y = element_blank(),
     axis.text.x = element_blank(), axis.ticks.x = element_blank(),
     axis.text.y = element_blank(), axis.ticks.y = element_blank(),
     legend.background = element_rect(fill = 'transparent', 
                                      colour = NA),
     legend.key = element_rect(fill = 'transparent', 
                               colour = NA),
     legend.position = "none", 
     legend.title = element_text(size=5), 
     legend.text = element_text(size=4),
     legend.key.height = unit(0.25, 'cm'),
     legend.key.width = unit(0.25, 'cm')
   )
)
```

```{r}
#| label: code-setup
set.seed(20240110)
source("nldr_code.R", local = TRUE)
```

@@

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(
  do.call(cbind, 
          gen_scaled_data(data = umap_pbmc, 
                          x = "UMAP1", 
                          y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_1 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.99.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_2 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.03.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_3 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.3, size = 3)

```

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_84_min_dist_0.99.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_4 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'e', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc_1 <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'f', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)   
```

```{r}
#| warning: false
#| echo: false
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'g', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.1, size = 3)   
```

```{r}
#| warning: false
#| echo: false
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_92.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc_2 <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'h', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)   
```

```{r}
#| warning: false
#| echo: false

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_2.rds")
phate_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = phate_pbmc, x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1", 
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(phate_pbmc))

phate_plot_pbmc_1 <- phate_pbmc_scaled |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'i', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
phate_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = phate_pbmc, x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1", 
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(phate_pbmc))

phate_plot_pbmc_2 <- phate_pbmc_scaled |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'j', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_10.rds")
phate_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = phate_pbmc, x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1", 
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(phate_pbmc))

phate_plot_pbmc_3 <- phate_pbmc_scaled |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'k', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
trimap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = trimap_pbmc, x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1",
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(trimap_pbmc))

trimap_plot_pbmc_1 <- trimap_pbmc_scaled |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'l', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_30_4_10.rds")
trimap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = trimap_pbmc, x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1",
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(trimap_pbmc))

trimap_plot_pbmc_2 <- trimap_pbmc_scaled |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'm', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_5_2_2.rds")
trimap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = trimap_pbmc, x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1",
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(trimap_pbmc))

trimap_plot_pbmc_3 <- trimap_pbmc_scaled |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'n', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_10_pca_0.5_2.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_1 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'p', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_2 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'q', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_5_pca_0.1_1.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_3 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'r', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.1_1.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_4 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 's', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```

```{r}
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Which is the most accurate representation of the structure in high dimensions?"
#| label: fig-diffnldrpbmc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
# 2D layouts from different NLDR techniques and different hyperparameter choices applied for the PBMC3k dataset: (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (d) UMAP (n_neighbors = 15, min_dist = 0.03), (e) UMAP (n_neighbors = 84, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (h) tSNE (perplexity = 92), (i) PHATE (knn = 2), (j) PHATE (knn = 5), (k) PHATE (knn = 10), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (m) TriMAP (n_inliers = 30, n_outliers = 4, n_random = 10), (n) TriMAP (n_inliers = 5, n_outliers = 2, n_random = 2), (p) PaCMAP (n_neighbors = 10, init = pca, MN_ratio = 0.5, FP_ratio = 2), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5), (r) PaCMAP (n_neighbors = 5, init = pca, MN_ratio = 0.1, FP_ratio = 1), and (s) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.1, FP_ratio = 1). Is there a best representation of the original data or are they all providing  equivalent information?
umap_plot_pbmc + umap_plot_pbmc_1 + umap_plot_pbmc_2 + #umap_plot_pbmc_3 + umap_plot_pbmc_4 + 
  tsne_plot_pbmc_1 + tsne_plot_pbmc + 
  #tsne_plot_pbmc_2 + phate_plot_pbmc_1 + 
  phate_plot_pbmc_2 + #phate_plot_pbmc_3 + 
  trimap_plot_pbmc_1 + #trimap_plot_pbmc_2 + trimap_plot_pbmc_3 + pacmap_plot_pbmc_1 + pacmap_plot_pbmc_2 + 
  pacmap_plot_pbmc_3 + #pacmap_plot_pbmc_4 + 
  plot_layout(ncol = 4)
```

### 2024-03-27 paper.tex
tidied up fist image and code

\begin{figure}[H]

@@

\caption{\label{fig-umapauthor}2D layout from UMAP (n\_neighbors = 30,
min\_dist = 0.3) applied for the PBMC3k dataset. Is this a best
representation of the original data?}

>>>>

\begin{figure}

@@

\caption{\label{fig-diffnldrpbmc}Six different NLDR representations of
the same data. Different techniques and different parameter choices are
used. Which is the most accurate representation of the structure in high
dimensions?}
